---
title: "Expression Quantification"
output: html_document
---
  
# Set up the environment

Define where the installed tools can be found. Adapt this to your installation
```{r}
STAR = "/usr/local/ngseq/src/STAR-STAR_2.4.2a/bin/Linux_x86_64/STAR"
SAMTOOLS = "/usr/local/ngseq/bin/samtools"
RSEM_PREPAPRE_REF = "/usr/local/ngseq/src/RSEM-1.2.22/rsem-prepare-reference"
RSEM_CALCULATE = "/usr/local/ngseq/src/RSEM-1.2.22/rsem-calculate-expression"
```

Define the directory where input and output data should be generated.
```{r, eval=TRUE}
dataDir = "/srv/GT/analysis/hubert/lecture/lecture-2015"
setwd(dataDir)
sampleNames = c("N1", "N2", "N3", "T1", "T2", "T3")
genomeFastaFile = "anno/genome.fa"
gtfFile = "anno/transcripts.gtf"
starIndexDir = "starIndex"
```

The data set consists for 3 Normal samples and 3 Treated samples with 2.5 Mio reads each. The reads are simulated to come from genes on the human chromosome 1. The file genome.fa contans that chromosome sequence. The file transcripts.gtf contaisn the gene coordinates.

```{r}
#list.files("FASTQ")
system(paste0("ls -lh ", dataDir, "/", "FASTQ"), intern = TRUE)
```


# Use STAR and featureCounts to count number of reads per gene

Create the STAR index
```{r, eval=FALSE}
dir.create(starIndexDir)
cmd = paste(STAR, " --runMode genomeGenerate",
 "--genomeDir", starIndexDir,
 "--genomeFastaFiles", genomeFastaFile,
 "--sjdbGTFfile", gtfFile, 
 "--runThreadN", 4)
system(cmd)
```

Define input and output files
```{r, eval=FALSE}
r1Files = paste0("FASTQ", "/", sampleNames, "_1.fq.gz")
r2Files = paste0("FASTQ", "/", sampleNames, "_2.fq.gz")
stopifnot(file.exists(r1Files))
stopifnot(file.exists(r2Files))
dir.create("STAR_mapped")
bamFiles = paste0("STAR_mapped", "/", sampleNames, ".bam")
```

Actual alignment run
```{r, eval=FALSE}
for (i in 1:length(sampleNames)){
  message(sampleNames[i])
  cmd = paste(STAR, "--genomeDir", starIndexDir,
              "--readFilesIn", r1Files[i], r2Files[i],
              "--readFilesCommand", "zcat",
              "--runThreadN", 16,
              "--outSAMtype BAM Unsorted")
  message(cmd)
  returnValue = system(cmd)
  stopifnot(returnValue == 0) ## only if returnValue is 0, the system cmd was successful
  cmd = paste(SAMTOOLS, "sort", "-m 2000M", "Aligned.out.bam", sub(".bam$", "", bamFiles[i]))
  returnValue = system(cmd)
  cmd = paste(SAMTOOLS, "index", bamFiles[i])
  returnValue = system(cmd)
}
```


# Create the featureCount table

We use the library subread to generate the counts
```{r, eval=FALSE}
dir.create("STAR_featureCounts")
library(Rsubread)

countResult1 = featureCounts(bamFiles, annot.ext=gtfFile, isGTFAnnotationFile=TRUE, isPairedEnd=TRUE, strandSpecific=0,
                                 GTF.featureType="exon", GTF.attrType="gene_id", useMetaFeatures=TRUE,
                                 allowMultiOverlap=TRUE, nthreads=16)
save(countResult1, file="countResult1.RData")
countResult2 = featureCounts(bamFiles, annot.ext=gtfFile, isGTFAnnotationFile=TRUE, isPairedEnd=TRUE, strandSpecific=0,
                                 GTF.featureType="exon", GTF.attrType="gene_id", useMetaFeatures=TRUE,
                                 allowMultiOverlap=TRUE, countMultiMappingReads=TRUE, fraction=TRUE, nthreads=16)
save(countResult2, file="countResult2.RData")
```


# Expression quantification with RSEM

Build the index for RSEM

```{r, eval=FALSE}
dir.create("rsemIndex")
rsemIndex = "rsemIndex/human"
cmd = paste(RSEM_PREPAPRE_REF,
            "--gtf", gtfFile,
            "--star", "--star-path", dirname(STAR),
            "-p", 8,
            genomeFastaFile,
            rsemIndex)
system(cmd)
```

Generate the expression estimates

```{r, eval=FALSE}
dir.create("RSEM_counts")

i = 1
countBasenames = paste0("RSEM_counts", "/", sampleNames)
for (i in 1:length(sampleNames)){
  message(sampleNames[i])
  cmd = paste(RSEM_CALCULATE,
              "--star", "--star-path", dirname(STAR),
              "-p", 16,
              "--calc-pme", "--calc-ci",
              "--gzipped-read-file",
              "--paired-end", r1Files[i], r2Files[i],
              rsemIndex,
              countBasenames[i])
  system(cmd)
}

```


## Exercises

1. Depending on the featureCount mode, how many alignments are ignored during counting?  
2. Compare the counts from featureCounts with the genelevel counts from RSEM. Describe the differences. Compute the correlation.  
3. How does the sum of the counts in RSEM compare to the number of the aligned reads?  
4. Compute pair-wise correlation of the samples based on the isoform counts and gene level counts

